Hijacking Simulators with Universal Probabilistic Programming

### 2019-05-22 paper/neurlps2019/main.tex

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{graphicx}
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{listings}
\usepackage{tabularx}
\usepackage{listings}
\usepackage{pgfplots}
\usepackage{tikz}
\usepackage{todonotes}
\usepackage{tikz-cd}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{wrapfig}
% \usepackage[binary-units=true]{siunitx}
\title{Hijacking Simulators with Universal Probabilistic Programming}

@@

\begin{wrapfigure}{R}{0.44\textwidth}
\centering
% \includegraphics[width=0.35\textwidth]{../plots/mpenny-pop10/mpenny_10.pdf}
\includegraphics[width=0.40\textwidth]{../plots/model_lambda/model_posterior.pdf}
\includegraphics[width=0.40\textwidth]{../plots/model_lambda/hist_posteriors.png}
\caption{An example of the posterior output of population-based simulator after hijacking. \textit{Top}: Represents the posterior execution traces of the simulator, inside the the probabilistic programming 
system. Each address $A_{i}$ correspond to a physical process within the simulator, we describe these addresses and their structures in more details in Section~\ref{sec:casestudy}.
\textit{Bottom}: Represents the histograms generated after posterior inference for the parameters of interest, in this case those parameters are represented by addresses $A_{2}$ to $A_{6}$, conditioned on some observations. 
The grey dashed lines represent the ground truth.}
\label{fig:hijacking} 
\end{wrapfigure}
One such machine learning paradigm that has the ability to resolve these issues is \emph{Probabilistic programming}\cite{gordon2014probabilistic,le2016inference}. 
Probabilistic programming enables users to write, condition on and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way. 
The source code of the given model, or equivalently the \emph{program}, is transformed into a program density, which can
be viewed as the joint density of the user-specified model, \emph{the simulator} in this instance.
Having direct access to the source code enables PPSs to conduct several processes automatically, such as 
the automation of complex statistical inference procedures\cite{zhou2019lf,goodman2012church,wood2014new}, which enables us to efficiently calibrate
model parameters, and we can also track how the internals of the program change during
run time, enabling a full analysis of the program structure and traces generated, which provides greater interpretability, see Figure~\ref{fig:hijacking}.
In addition to this, as models are directly expressed in program code we can encode other structures to make use of the full structure of a given language, whether that is \texttt{python, C++} or 
any other language,
which enables users to integrate PPSs into their standard model development process, increasing efficiency and reducing model iteration time.

For all the brilliance of probabilistic programming, if one did want to run an existing simulator in a standard probabilistic programming set-up one would have to entirely re-write the simulator in the language of the PPS.
This in many cases is not feasible, nor efficient due to both the size 
and complexity of the existing simulator code base. 
However, what if one could connect arbitrary a simulator, written in any
language and directly connect it to a PPS, without having to re-write
the simulator in the given PPS?

\paragraph{Related work} Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient}. 
They demonstrated how this could be done for event-based simulators, but there methods neglected a large class of population-based simulators that model interactions 
between populations, that are utilized for modeling epidemics, multi-agent systems and financial economics  \bg{add citations}. 
Our framework provides a solution and makes it easy to transform arbitrary stochastic population-based simulator, in addition to the event-based simulators, into probabilistic programs, regardless of the complexity of the simulator. 
Our system supports simulators written in 13 different programming languages and enables the user to perform posterior inference over the density of the simulator, conditioned on what data the user has access too, including 
simulator output. We provide a theoretical justification for our approach and many examples.
% This enables two things.
% 1) It enables software developers and researchers to understand complicated code bases. 
% 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user of the inference results understands what physical events led to inference outcome. 
% This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.

@@

%Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.

@@

This paper first gives an overview of existing Malaria simulators (Section \ref{sec:background}), and proceed by introducing the necessary background on the \textit{pyprob} framework and the concept of universal probabilistic programming (Section \ref{sec:hijackingsimulators}). Our approach is then demonstrated and analysed in the context of a Malaria case study (Section \ref{sec:casestudy}).

@@

 \label{sec:methods}

  \begin{algorithm}
   \caption{How to perform posterior inference over a stochastic simulator with 
   }
    \begin{algorithmic}[1]
      \Function{Convert Simulator to a Probabilistic Program}{$A$}\Comment{}

        \State replace raw samples statements $f_{t}(x_{t} | \eta_{t}) $
        \State Create thin wrapper
        \State Create Containerized Environment $C_{i}$
        \State Create Connection
        \State Run Program
       \EndFunction
       \Function{Create thin wrapper}{$f_{t}(x_{t} | \eta_{t})$}
       \State to add
       \EndFunction
       \Function{Create Containerized Environment }{$\Delta_{i}$}\Comment{Where $\Delta_{i}$ is one of $i$-th dependencies}
       \State to add
       \EndFunction

@@

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{graphicx}
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{listings}
\usepackage{tabularx}
\usepackage{listings}
\usepackage{pgfplots}
\usepackage{tikz}
\usepackage{todonotes}
\usepackage{tikz-cd}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{wrapfig}

\newcommand{\n}[1]{\ensuremath{n_{#1}}}
\newcommand{\rv}[1]{\ensuremath{{#1}_{1:n_{#1}}}}
\newcommand{\rvv}[1]{\ensuremath{{#1}_{1:n_{#1'}}}}
\newcommand{\model}[1]{\ensuremath{\mathcal{M}{(#1)}}}
\newcommand{\simu}[1]{\ensuremath{\Omega(\model{#1})}}
\newcommand{\micro}[2]{\ensuremath{\phi_{#1}(#2)}}
\newcommand{\pdf}[1]{\ensuremath{\mathcal{P}_{d}(#1)}}
\newcommand{\pmf}[1]{\ensuremath{\mathcal{P}_{m}(#1)}}
\newcommand{\add}[1]{\ensuremath{A_{#1}}}
\newcommand{\adds}[2]{\ensuremath{A_{#1}_{#2}}}
\newcommand{\expect}[1]{\ensuremath{\mathbb{E}[#1]}}
\newcommand{\trace}[1]{\ensuremath{\mathcal{T}_{#1}}}
\newcommand{\traces}[2]{\ensuremath{\textit{tr}^{#1}_{#2}}}
\newcommand{\stoc}[2]{\ensuremath{f^{#1}_{#2}(\rvv{x})}}
% \usepackage[binary-units=true]{siunitx}
\title{Hijacking Simulators with Universal Probabilistic Programming}

@@

One such machine learning paradigm that has the ability to resolve these issues is \emph{Probabilistic programming}\cite{gordon2014probabilistic,milch2005blog,van2018introduction,tran2017deep}. 
Probabilistic programming enables users to write, condition on and perform inference in, complex probabilistic models in a programmatic, yet efficient and simplified way. 
The source code of the given model, or equivalently the \emph{program}, is transformed into a program density, which can
be viewed as the joint density of the user-specified model, \emph{the simulator} in this instance.
Having direct access to the source code enables PPSs to conduct several processes automatically, such as 
the automation of complex statistical inference procedures\cite{zhou2019lf,goodman2012church,wood2014new,rainforth2017automating}, which enables us to efficiently calibrate
model parameters, and we can also track how the internals of the program change during
run time, enabling a full analysis of the program structure and traces generated\cite{milch2005blog}, which provides greater interpretability, see Figure~\ref{fig:hijacking}.
In addition to this, as models are directly expressed in program code we can encode other structures to make use of the full structure of a given language, whether that is \texttt{python, C++} or 
any other language,
which enables users to integrate PPSs into their standard model development process, increasing efficiency and reducing model iteration time.

\begin{wrapfigure}{R}{0.44\textwidth}
\vspace{-20pt}
\centering
% \includegraphics[width=0.35\textwidth]{../plots/mpenny-pop10/mpenny_10.pdf}
\includegraphics[width=0.40\textwidth]{../plots/model_lambda/model_posterior.pdf}
\includegraphics[width=0.40\textwidth]{../plots/model_lambda/hist_posteriors.png}
% \vspace{-10pt}
\caption{An example of the posterior output of population-based simulator after hijacking. \textit{Top}: Represents the posterior execution traces of the simulator, inside the the probabilistic programming 
system. Each address $A_{i}$ correspond to a physical process within the simulator, we describe these addresses and their structures in more details in Section~\ref{sec:casestudy}.
\textit{Bottom}: Represents the histograms generated after posterior inference for the parameters of interest, in this case those parameters are represented by addresses $A_{2}$ to $A_{6}$, conditioned on some observations. 
The grey dashed lines represent the ground truth.}
\vspace{-25pt}
\label{fig:hijacking} 
\end{wrapfigure}
For all the brilliance of probabilistic programming, if one did want to run an existing simulator in a standard probabilistic programming set-up one would have to entirely re-write the simulator in the language of the PPS.
This in many cases is not feasible, nor efficient due to both the size 
and complexity of the existing simulator code base. 
However, what if one could connect arbitrary a simulator, written in any
language and directly connect it to a PPS, without having to re-write
the simulator in the given PPS?

\paragraph{Related work} Our method of hijacking simulators builds on the work of~\cite{baydin2018efficient,casado2017improvements}. 
They demonstrated how this could be done for event-based simulators, but there methods neglected a large class of population-based simulators that model interactions

between populations, that are utilized for modeling epidemics, multi-agent systems and financial economics \cite{raberto2001agent,landau_binder_2014,hartig2011statistical,smith2008towards,bershteyn2018implementation}. 
Our framework provides a solution and makes it easy to transform arbitrary stochastic population-based simulator, in addition to the event-based simulators, into probabilistic programs, regardless of the complexity of the simulator. 
Our system supports simulators written in 13 different programming languages and enables the user to perform posterior inference over the density of the simulator, conditioned on what data the user has access too, including 
simulator output. 
We provide a theoretical justification for our approach and many examples.
% This enables two things.
% 1) It enables software developers and researchers to understand complicated code bases. 
% 2) It provides interpretable inference results that provide policy makers with predictions that are truly interpretable, in the sense that the end-user, of the inference results understands what physical events led to inference outcome. 
% This is critical in several domains, and is particularly important in the medical domain. We provide examples of both in Section~\textbf{TO ADD REF}.

@@

% Moreover, though our focus is on hijacking epidemiology simulators, our general approach and framework can be applied more widely to a general class of stochastic simulators.

@@

@@

\begin{center}
\begin{tabular}{ |c|c|c| }
\hline
Symbol & Meaning  \\
\hline
\n{x} & Number of latent variables \\
\n{y} & Number of input observations \\
\n{y’} & Number of output observations, generated simulator values \\
\rv{x} & The latent parameters \\
\rv{y} & The input observed parameters \\
\rv{y’} & The output observed parameters \\
\model{\cdot} & Model - i.e .xml file \\
\simu{\cdot} & Simulator \\
\micro{k}{\cdot} & $k$-th Micro-simulation \\
\pdf{\cdot} & probability density function \\
\pmf{\cdot} & probability mass function \\
\add{i} & Address $i$ \\
\expect{\cdot} & Expectation (first moment) \\
\trace{j} & The $j$-th collapsed Trace \\
\traces{i}{j} & A single, un-collapsed Trace for the $i$th trace. ``A program trace’' \\
\stoc{h}{i} & A stochastic process at step $h$ for \add{i} \\
\hline
\end{tabular}
\end{center}
  \begin{algorithm}
  \small
   \caption{Hijacking a stochastic simulator for an arbitrary machine}
    \begin{algorithmic}\small
      \Function{Simulator to a Probabilistic Program}{Simulator $\Omega(\cdot)$, Model input to simulator \model{\cdot}, PPS sampling primitive $f_t()$, Message passing protocol (PPX) $h_t()$, Random variable $x_{sim}$, PPS $\mathcal{P}$ }
       \State  $ x_{sim} \rightarrow f_{t}(x_{sim}) $\Comment{Replace raw sample statements}
       \State $\simu{f_{t}(x^{t}_{sim})} \rightarrow  h_{t}(\simu{f_{t}(x^{t}_{sim})}) \rightarrow \mathcal{P}(f_{t}(x^{t}_{sim}))$ \Comment{An API between the simulator enables raw RNG to be passed through the PPS}

        \Function{Create Containerized Environment}{$C(\Delta)$}
        \For{$\delta$ in $\Delta$}
        \State Add dependency $\delta$
        \EndFor
        \State Run Program
       \EndFunction
       \Function{Create thin wrapper}{$f_{t}(x_{t} | \eta_{t})$}
       \State to add
       \EndFunction
       \Function{Create Containerized Environment }{$\Delta_{i}$}\Comment{Where $\Delta_{i}$ is one of $i$-th dependencies}
       \State to add
       \EndFunction
       \EndFunction

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{../plots/mastsari/mastsari_prior_trace_structure.pdf}
  \caption{Prior trace probabilities of the Mastsari model.}
  \label{fig:priortracemastsari}
\end{figure*}
